\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-24 paper.tex

\section{Computational efficiency and data interchange}
\label{sec-efficiency}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of possible applications
% TODO get rid of this, it's a duplication of the introduction
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely,ignatieva2023distribution}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and has subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
(Technically, edges in tree sequence terminology would be better
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes. This
denormalisation of the gARG data model in \texttt{tskit} is for efficiency purposes.)
The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2022slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}.

% FIXME this needs more work and doesn't really fit into the
% narrative right now, but we do need to make this point.
The key insight that makes the succinct tree sequence encoding
an efficient substrate for defining analysis algorithms is that
it allows us to recover the local trees along the genome
in a way that allows us to reason about the \emph{differences}
between those trees.
Sequentially generating the local trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While this has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For instance, several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures, which often involve overlapping generations and extreme variation in lifetime reproductive success. Flexible ARG encoding should accommodate such cases.

@@

Interest in ARG inference methods and downstream applications is
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
Here, we have advocated
for decoupling the data structures used to represent ARGs
from generative processes, and have shown that the
gARG encoding enables flexible description of
contemporary datasets with varying levels of precision and integral
handling of some important forms of temporal uncertainty.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem.
Ideally, this would comprise a wide range of different
inference methods specialised to different organisms, types
and scales of data and
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to
% get good local trees
specific inference goals, sharing a common ARG interchange format.
The output of these diverse inference methods could then be
processed by numerous downstream applications,
enabling users to freely compare the outputs of different
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared
this vision,
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Today, however, we have the benefit of a pre-existing ARG
interchange format, already widely used and accompanied
by high-quality, high-performance software
(see Section~\ref{sec-efficiency} for details).
Adopting the \texttt{tskit} library
as a community standard,
utilising its rich suite of operations,
may accelerate progress and help to
establish ARG methods within mainstream genetics.

>>>>

\section{Implementation and efficiency}
\label{sec-efficiency}
The gARG encoding discussed defined here
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure
(usually known as a ``tree sequence'' for brevity)
is a practical gARG implementation focused on efficiency.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and has subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
inference from data~\citep{kelleher2019inferring,wohns2022unified},
and calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The key property of the succinct tree sequence encoding
that makes it an efficient substrate for defining analysis
algorithms is that it allows us to sequentially
recover the local trees along the genome very efficiently,
and in a way that allows us to reason about the \emph{differences}
between those trees~\citep{kelleher2016efficient,ralph2020efficiently}.

The \texttt{tskit} library is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,
guo2022recombination,
korfmann2022weak,
mahmoudi2022bayesian,
petr2022slendr,
rasmussen2022espalier,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}.
The technical details of \texttt{tskit}, and how it provides an
efficient and portable platform for ARG-based analysis, are beyond
the scope of this manuscript.
In the interest of avoiding confusion, however, we list a
few minor details in which the formal details of gARGs
provided here differ from their practical implementation in
\texttt{tskit}.
Firstly, ``edges'' in tree sequence terminology would be perhaps better be
described as ``edge-intervals'', as each describes a single contiguous
interval of genome inheritance between a pair of nodes.
This denormalisation of the gARG data model in is for efficiency purposes.
Secondly, $-1$ is used to denote the presence of
roots in an oriented tree, rather than $0$ as used here for notational
simplicity.

% Other representations of ARGs, beyond straightforward implementations
% of a Griffiths-style
% eARG~\citep[e.g.][]{ignatieva2021kwarg,rasmussen2014genome}
% and \texttt{tskit}'s gARG format, are of course possible.
% The ARGON simulator~\citep{palamara2016argon}, for example,
% stores ARGs
% ARG-Needle~\citep{zhan2023towards}, for example, uses a format based
% on hap

@@

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While this has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For instance, several human datasets now consist of hundreds of thousands of
genomes~\citep{turnbull2018hundred, bycroft2018genome,
karczewski2020mutational,tanjo2021practical,
halldorsson2022sequences},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures, which often involve overlapping generations and extreme variation in lifetime reproductive success. Flexible ARG encoding should accommodate such cases.

@@

Interest in ARG inference methods and downstream applications is
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem.
Ideally, this would comprise a wide range of different
inference methods specialised to different organisms, types
and scales of data and
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to
% get good local trees
specific inference goals, sharing a common ARG interchange format.
The output of these diverse inference methods could then be
processed by numerous downstream applications,
enabling users to freely compare the outputs of different
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared this vision,
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems with converting data from one program's
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
Current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
which is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse software ecosystem.
The gARG encoding described here is a significant generalisation
of classical concepts, capable of describing even the bewildering complexity
of contemporary datasets and describing important forms of temporal uncertainty.
The \texttt{tskit} library is a high-quality open-source gARG implementation,
with proven efficiency and
scalability~\citep[e.g.][]{anderson2022genes,zhan2023towards},
that is already in widespread use.
Adopting it as a community standard,
and utilising its rich suite of operations,
may accelerate progress and help to
establish ARG methods within mainstream genetics.
